CrypTool Script \\ Mathematics and Cryptography

### 2002-08-26 script/en/chapters/primes.tex
be comments

The 37th number was found in January 1998 and has 909,526
digits in the decimal system, which corresponds to 33 pages in the newspaper!

\newpage
These numbers can be found at the following URLs:

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-38 -- June 1999}\index{Mersenne!prime number!M-38}
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

@@

\subsubsection{Prime number tests}
\index{Prime number!test}
In order to implement secure encryption procedures we need extremely large prime
numbers (in the region of $2^{2,048}$, i.e. numbers with $600$ digits in the
decimal system!).

@@

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Pseudoprime numbers}\index{Number!pseudoprime} {\em pseudoprime numbers}. The first pseudoprime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Number!Carmichael} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudoprime numbers}. Again, there are
strong pseudoprime numbers that are not primes, but this is much less often the
case than for (simple) pseudoprime numbers. The smallest strong pseudoprime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudoprime number up to $25 \cdot 10^9$, i.e. a number that passes the test and
yet is not a prime number.

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful. Comment: However, what is special about all these numbers
is that any pair of them does not have a common factor other than $1$. They are
therefore \index{Prime number!relative} {\em relatively prime}.
\pagebreak
  \item $f(n) = n^2 + n + 41$:\\
  This sequence starts off very {\em promisingly},   but is far from being a
proof.

@@

    There exists no such polynomial that for all $x$ in ${\mathbb Z}$ only
delivers prime values.
    For a proof of this, please refer to \cite[p. 83 f.]{Padberg1996}, where you
will also find further details about prime number formulae.
  \item  \index{Catalan numbers} Catalan, after whom the so-called
{\em Catalan numbers} $A(n) = (1 / (n+1) ) * (2n)! / (n!)^2$ are named,
conjectured that $C_4$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{http://www.math.Pr
inceton.EDU/\~{}arbooker/nthprime.html}.} \vskip +7 pt Output of the
$n$-th prime number

@@

@@

\begin{enumerate}
   \item \href{http://www.utm.edu/}{\tt http://www.utm.edu/ }
   \item
\href{http://prothsearch.net/index.html}{\tt
http://prothsearch.net/index.html}
   \item \href{http://www.mersenne.org/prime.htm}{\tt
http://www.mersenne.org/prime.htm }
   \item \href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt
http://reality.sgi.com/chongo/prime/prime\_press.html }
   \item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt
http://www.eff.org/coop-awards/prime-release1.html }
\item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt
%http://www.octocad.demon.co.uk/mersenne/prime.htm }
%   \item \href{http://www.informatik.tu-darmstadt.de/TI/LiDIA/}{\tt
http://www.informatik.tu-darmstadt.de/TI/LiDIA/ }
   \item \href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{\tt
http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html }
   \item \href{http://www.utm.edu/research/primes/notes/by_year.html}{\tt
http://www.utm.edu/research/primes/notes/by\_year.html }
   \item \href{http://www.cerias.purdue.edu/homes/ssw/cun}{\tt
http://www.cerias.purdue.edu/homes/ssw/cun }
        \item \href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
        \item \href{http://www.mathematik.ch/mathematiker/goedel.html}{http://www.mathematik.ch/mathematiker/goedel.html}

>>>>

These numbers can be found at the following URLs:

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-37 -- January 1998}\index{Mersenne!prime number!M-37}
The 37th Mersenne prime,
$$ 2^{3.021.377} - 1. $$
was found in January 1998 and has 909,526
digits in the decimal system, which corresponds to 33 pages in the newspaper!

\subsubsection{M-38 -- June 1999}\index{Mersenne!prime number!M-38}
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

@@

\subsection{Prime number tests}
\index{Prime number!test}
In order to implement secure encryption procedures we need extremely large prime
numbers (in the region of $2^{2,048}$, i.e. numbers with $600$ digits in the
decimal system!).

@@

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Pseudoprime number}\index{Number!pseudoprime} {\em pseudoprime numbers}. The first pseudoprime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Number!Carmichael} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudoprime numbers}. Again, there are
strong pseudoprime numbers that are not primes, but this is much less often the
case than for (simple) pseudoprime numbers. The smallest strong pseudoprime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudoprime number up to $25 \cdot 10^9$, i.e. a number that passes the test and
yet is not a prime number.

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful. Comment: However, what is special about all these numbers
is that any pair of them does not have a common factor other than $1$. They are
therefore \index{Prime number!relative} {\em relatively prime}.

  \item $f(n) = n^2 + n + 41$:\\
  This sequence starts off very {\em promisingly},   but is far from being a
proof.

@@

    There exists no such polynomial that for all $x$ in ${\mathbb Z}$ only
delivers prime values.
    For a proof of this, please refer to \cite[p. 83 f.]{Padberg1996}, where you
will also find further details about prime number formulae.
\pagebreak
  \item  \index{Catalan number} Catalan, after whom the so-called
{\em Catalan numbers} $A(n) = (1 / (n+1) ) * (2n)! / (n!)^2$ are named,
conjectured that $C_4$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html}.} \vskip +7 pt Output of the
$n$-th prime number

@@

\bibitem[Doxiadis2000]{Dioxadis2000}
    Apostolos Doxiadis, \\
    Uncle Petros and the Goldbach's Conjecture, Faber/Bloomsbury, 2000.

@@

\begin{enumerate}
   \item \href{http://www.utm.edu/}{\tt http://www.utm.edu/ }
   \item
\href{http://prothsearch.net/index.html}{\tt
http://prothsearch.net/index.html}
   \item \href{http://www.mersenne.org/prime.htm}{\tt
http://www.mersenne.org/prime.htm }
   \item \href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt
http://reality.sgi.com/chongo/prime/prime\_press.html }
   \item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt
http://www.eff.org/coop-awards/prime-release1.html }
\item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt
%http://www.octocad.demon.co.uk/mersenne/prime.htm }
%   \item \href{http://www.informatik.tu-darmstadt.de/TI/LiDIA/}{\tt
http://www.informatik.tu-darmstadt.de/TI/LiDIA/ }
   \item \href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{\tt
http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html }
   \item \href{http://www.utm.edu/research/primes/notes/by_year.html}{\tt
http://www.utm.edu/research/primes/notes/by\_year.html }
   \item \href{http://www.cerias.purdue.edu/homes/ssw/cun}{\tt
http://www.cerias.purdue.edu/homes/ssw/cun }
        \item \href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
        \item \href{http://www.mathematik.ch/mathematiker/goedel.html}{\tt http://www.mathematik.ch/mathematiker/goedel.html}
        \item \href{http://www.mscs.dal.ca/~dilcher/golbach/index.html}{\tt http://www.mscs.dal.ca/\~{}dilcher/goldbach/index.html}
